---
title: "Mortality in NAFLD"
author: "Ian Rowe"
date: "05/09/2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r code, include = FALSE}

library(tidyverse)
library(pander)

MainCauses <- read_csv("MainCauses.csv")

Table1 <-
  MainCauses %>%
  rename(Cause = Group1,
         Percentage = Proportion,
         Number = n) %>%
  select(-Total) %>%
  filter(Percentage >2.1)



data2018CoDNAFLD <- read_csv("data2018CoDNAFLD.csv")

Table2 <-
  data2018CoDNAFLD %>%
  rename(Cause = Group1,
         Percentage = Proportion,
         Number = n) %>%
  select(-Total) %>%
  filter(Percentage >1)

```

## How often do patients with NAFLD die from liver disease?
Data presented from Dr Younossi's group at the recent ILC challenged the view that cardiovascular disease and extrahepatic cancer are the commonest causes of death in patients with NAFLD (Abstract AS158).

The study used National Vital Statsitics data that are available at https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm.  Exploring this reveals important insights about the capture of patients with NAFLD.  The outputs are not identical to those shown in the presentation, likely due to analytical decisions.  All code for these analyses is available at https://github.com/iacrowe/NAFLD-mortality.


## Causes of death in the US 2018
There were 2.8 million deaths in 2018.  The commonest causes of death were cardiovascular and cancer, accounting for 30% and 20% of all deaths respectively.  Liver related mortality was ranked 10th, accounting for 2% of all deaths (Table 1).

```{r Table 1, echo = FALSE}

panderOptions('table.alignment.default', "left")

pander(Table1)

```
#### Table 1. Top 10 causes of death in the US in 2018.

## Causes of death in NAFLD
Using the same definition for NAFLD and similar methods 39,434 individuals with NAFLD were identified in the death certificate dataset.  That is 1.3% of all deaths.  Among that group, 55% of all deaths in patients identified with NAFLD who died in 2018 were attributable to complications of liver disease.  A further 5% were attributed to HCC (Table 2).

This substantiates the findings of the study presented at the ILC, but it highlights the citical issue of the definition of the denominator.

```{r Table 2, echo = FALSE}

panderOptions('table.alignment.default', "left")

pander(Table2)

```
#### Table 2. Top 10 causes of death in decedents identified with NAFLD.


## Projecting causes of death to all those with NAFLD
A conservative estimate of the prevalence if NAFLD amongst those dying in 2018 is 25% given current estimates of the population prevalence rather than the 1.3% identified in death certificate data.  Since liver failure is readily diagnosed and coded, it is likely that the vast majority of cases are recorded in the registry.  

Using the population prevalence estimate as the denominator substantially changes the estimate of mortality from liver disease, reducing from 55% to 3.5%.

## Liver disease remains an uncommon cause of death in individuals in NAFLD
The existing literature supports cardiovascular disease and extrahepatic cancer as the commonest causes of death in NAFLD.  The limitations in the dataset employed in this recent study presented at the ILC should not change this view.


